Differentiable transit models for ellipsoidal planets

### 2024-05-23 src/tex/ms.tex
[showyourwork] overleaf sync

% Title
\title{Semi-analytic transit models for oblate planets in Jax}

@@

% Main body with filler text
\section{Introduction}
\label{sec:intro}

@@

In the elliptical case, we can modify Eq.~\ref{eq:circularplanet} by deforming the occultor along the y axis by a value $b$, now taking the radius $r_o$ to represent the projected equatorial radius of the occultor. 
\begin{equation} \label{eq:ellipticalplanet}
(x-x_{o})^2-\frac{(y-y_o)^2}{b^2} = r_o^2
\end{equation}
where we define $b=1-f$. Here we emphasize that the $r_o$ and $f$ refer to the \textit{projected} equatorial radius and oblateness respectively. 
Solving for y in the above equation and then plugging it into Eq.~\ref{eq:unitcircle} yields a quartic polynomial of the form: 
\begin{equation} \label{eq:quarticform}
Ax^4 + Bx^3 + Cx^2 + Dx + E = 0
\end{equation} 
where
\begin{align*}
A &= \frac{b^4 - 2b^2 + 1}{4y_o^2}\\
B &= \frac{-b^4x_o + b^2x_o}{y_o^2}\\
C &= \frac{-b^4r_o^2 + 3b^4x_o^2 + b^2r_o^2 - b^2x_o^2 + b^2y_o^2 + b^2 + y_o^2 - 1}{2y_o^2} \\
D &= \frac{b^4r_o^2x_o - b^4x_o^3 - b^2x_oy_o^2 - b^2x_o}{y_o^2} \\
E &= \frac{b^4r_o^4 - 2b^4r_o^2x_o^2 + b^4x_o^4 - 2b^2r_o^2y_o^2 - 2b^2r_o^2 + 2b^2x_o^2y_o^2 + 2b^2x_o^2 + y_o^4 - 2y_o^2 + 1}{4y_o^2}\\
\end{align*}
We note that other methods of finding the flux in transit of elliptical occultors also require solving for quartic polynomials \citep{rein2023}. 
\subsection{Planet boundary integral}

\begin{figure}[ht!]
    \script{integral_bounds.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/oblate_planet.pdf}
        \caption{Geometry of the problem of computing the flux due to an oblate occultor as presented in this paper. 
        }
        \label{fig:integral_bounds}
    \end{centering}
\end{figure}

\subsection{Limb darkening}
\subsection{Oblate Case}
\subsection{Prolate Case}
\section{Prospects for Detection with JWST and other Instruments}
\label{sec:jwstdetect}
\subsection{Cramer-Rao Recoverability Estimates}
\label{sec:rec}
\subsection{Fisher Forecasting of Existing Planet Population}
\label{sec:fisherforecasting}
\subsection{Impact of Limb Darkening}
\subsection{Impact on Atmospheric Retreivals}
\section{Implementation in \lowercase{\texttt{jax0planet}}}
\label{sec:jax0planet}
\section{Application to [test system]}
\label{sec:lctest}
\section{Discussion and Future Work}
\label{disc}
\subsection{Hierarchical Inference}
\bibliography{bib}

>>>>

% Title
\title{Differentiable transit models for ellipsoidal planets}

@@

% Main body with filler text
\section{Introduction}
The contemporary study of exoplanets frequently assumes spherical bodies. While this is often a valid assumption, the deviations from this case are informative and contain valuable information which may be discarded by assuming sphericity.
\label{sec:intro}

@@

In the elliptical case, we can modify Eq.~\ref{eq:circularplanet} by deforming the occultor along the y axis by a value $b$, now taking the radius $r_o$ to represent the projected equatorial radius of the occultor. 
\begin{equation} \label{eq:ellipticalplanet}
(x-x_{o})^2-\frac{(y-y_o)^2}{b^2} = r_o^2
\end{equation}
where we define $b=1-f$. Here we emphasize that the $r_o$ and $f$ refer to the \textit{projected} equatorial radius and oblateness respectively. 
Solving for y in the above equation and then plugging it into Eq.~\ref{eq:unitcircle} yields a quartic polynomial of the form: 
\begin{equation} \label{eq:quarticform}
Ax^4 + Bx^3 + Cx^2 + Dx + E = 0
\end{equation} 
where
\onecolumngrid
\begin{equation} \label{eq:quarticcoeffs}
\begin{aligned}
A &= \frac{b^4 - 2b^2 + 1}{4y_o^2}\\
B &= \frac{-b^4x_o + b^2x_o}{y_o^2}\\
C &= \frac{-b^4r_o^2 + 3b^4x_o^2 + b^2r_o^2 - b^2x_o^2 + b^2y_o^2 + b^2 + y_o^2 - 1}{2y_o^2} \\
D &= \frac{b^4r_o^2x_o - b^4x_o^3 - b^2x_oy_o^2 - b^2x_o}{y_o^2} \\
E &= \frac{b^4r_o^4 - 2b^4r_o^2x_o^2 + b^4x_o^4 - 2b^2r_o^2y_o^2 - 2b^2r_o^2 + 2b^2x_o^2y_o^2 + 2b^2x_o^2 + y_o^4 - 2y_o^2 + 1}{4y_o^2}\\
\end{aligned}
\end{equation}

\twocolumngrid
We note that other methods of finding the flux in transit of elliptical occultors also require solving for quartic polynomials \citep{rein2023}. We solve the roots of this polynomial using eigendecomposition of the companion matrix (see \ref{sec:jax0planet} for details of the implementation). \textcolor{red}{we need to update this in eclipsoid to use a more numerically precise method}. The solution gives the x values of the intersection points where the projected disk of the star and the occultor coincide. We can then find the corresponding y values by plugging it back into the formula for either the occultor or occulted body.

\subsection{Star boundary integral}
The first integral $\mathcal{Q}(\mathbf{G}_n)$ is performed around the boundary of the occulted body's projected disk (bolded black border in Fig.~\ref{fig:integral_bounds}.) While the integrand is the same as in \citet{starry2019}, the bounds of the integral are the roots of the quartic polynomial shown in Eq.~\ref{eq:quarticcoeffs}. We compute the angle $\xi$, defined as the angle between the x-axis and a given intersection point, for all the intersection points. We then sort these angles in clockwise order.

\subsection{Occultor boundary integral}
The occultor (or planet) boundary integral $\mathcal{P}(\mathbf{G}_n)$ also starts with the intersection points. We define an angle $\phi$ to parametrize the bounds of the line integral. This angle, as noted in \citet{dholakia2022}, is defined similarly to an eccentric anomaly; the angle from the semimajor axis of the planet to the perpendicular projection of an intersection point onto the circle bounding the ellipse. Then, for the integrand, we start with the parametric formula for an ellipse:

\begin{align}
    x &= r_o \cos(\phi) + x_o \\
    y &= r_o b \sin(\phi) + y_o
\end{align}
We then plug this into the integrand for $\mathcal{P}(\mathbf{G}_n)$ in Eq.~\ref{eq:pandq} to obtain:
\begin{align}
\mathcal{P}(\mathbf{G}_n) = \int_{\phi}^{2\pi + \phi}[\ G_{ny}(r_o c_\phi + x_o , r_o b s_\phi + y_o) b c_\phi] \ r_o d\phi \\
- \int_{\phi}^{2\pi + \phi}[G_{nx}(r_o c_\phi + x_o, r_o b s_\phi + y_o)s_\phi]\ r_o d\phi
\end{align}
where we write $\sin{(\phi)}$ as $s_\phi$ and $\cos{(\phi)}$ as $c_\phi$ for brevity.
\begin{figure}[ht!]
    \script{integral_bounds.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/oblate_planet.pdf}
        \caption{Geometry of the problem of computing the flux due to an oblate occultor as presented in this paper. 
        }
        \label{fig:integral_bounds}
    \end{centering}
\end{figure}

\subsection{Limb darkening}
\subsection{Oblate Case}
\subsection{Prolate Case}
\section{Implementation in \lowercase{\texttt{jax0planet}}} \label{sec:jax0planet}
-implementation details:\\
    -API\\
    -root finding (with custom derivatives)\\
    -gauss quad\\
    -plot showing it matches brute force for each term in Gn   \\
\begin{figure}[ht!]
    \script{bruteforce_comparison.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/bruteforce_comparison.pdf}
        \caption{Comparison of the 1D Green's integral to a brute force 2D integral for each term in Green's basis. Residuals are under the expected numerical error in the brute force solution for all terms.
        }
        \label{fig:bruteforce_comparison}
    \end{centering}
\end{figure}

-speed\\
-precision (do we reach machine precision?)\\
-show derivatives, maybe against numerical diff\\
\section{Prospects for Detection with JWST and other Instruments}
\label{sec:jwstdetect}

\subsection{Cramer-Rao Recoverability Estimates}
\label{sec:rec}

\subsection{Fisher Forecasting of Existing Planet Population}
\label{sec:fisherforecasting}
\subsection{Impact of Limb Darkening}
\section{Application to [test system]}
\label{sec:lctest}
\section{Discussion and Future Work}
\label{disc}
\subsection{Hierarchical Inference}
\bibliography{bib}
